---
title: 'Lab 1: Question 2'
author: "Yao Chen, Jenny Conde, Satheesh Joseph, Paco Valdez, Yi Zhang"
output: pdf_document
---

```{r load packages, echo=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2) 
# library(tidyverse) # if you want more, but still core, toolkit
```

```{r load and clean data, echo=FALSE, warning=TRUE, message=FALSE}
# Load data
rawData <- read_dta('data/anes_timeseries_2020_stata_20210211.dta')

# Useful variables
# V201018 -- PRE: PARTY OF REGISTRATION
# V201151 -- PRE: FEELING THERMOMETER: JOE BIDEN, DEMOCRATIC PRESIDENTIAL CANDIDATE
# V201153 -- PRE: FEELING THERMOMETER: KAMALA HARRIS, DEMOCRATIC VICE-PRESIDENTIAL CANDIDATE
highSentimentThreshold = 100
lowSentimentThreshold = 0
DemocratPartyCode = 1

democraticDf <- data.frame(party=rawData$V201018, Biden=rawData$V201151, Harris = rawData$V201153) %>% 
  filter(party == DemocratPartyCode)
democraticDf <- within(democraticDf, Harris[Harris > highSentimentThreshold] <- highSentimentThreshold)
democraticDf <- within(democraticDf, Biden[Biden > highSentimentThreshold] <- highSentimentThreshold)
democraticDf <- within(democraticDf, Harris[Harris < lowSentimentThreshold] <- lowSentimentThreshold)
democraticDf <- within(democraticDf, Biden[Biden < lowSentimentThreshold] <- lowSentimentThreshold)
democraticDf[order(democraticDf$Biden),]

cat(" Biden : Mean - ", mean(democraticDf[["Biden"]]), " sd - ", sd(democraticDf[["Biden"]]), "\n")
cat(" Harris : Mean - ", mean(democraticDf[["Harris"]]), " sd - ", sd(democraticDf[["Harris"]]), "\n")
```


```

rawData
```

## Importance and Context
The recent election results are path breaking.
It yielded, The first female vice president; The first Asian-American vice president. 
The United States of America embarked on a new path. 
A path made of equality. A path made of diversity. A path made of principles.
I am fascinated to see the first of its kind combination - Biden and Harris.
Like everyone, I see both of them have the substance and charisma to lead the nation.

But, as a Data Scientist, I found it intriguing to explore the enthusiasm of democratic voters between them. 
Thus, the hypothesis.
Are Democratic voters more enthusiastic about Joe Biden or Kamala Harris.

## Description of Data
<!-- Explain how your research question is operationalized, including whether the variables you create are appropriate to study the concepts in question. --> 
The American National Election Stuides (ANES) conducts surveys of voters in the United States, with a flagship survey occurring immediately before and after each presidential election. 
Our research question is 

<!-- What are some basic features of the data distribution? --> 
<!-- What changes do you make to the data and why? --> 
```{r desc, echo=FALSE, warning=TRUE, message=FALSE}

```

## Most appropriate test 
<!-- Explain which test is the most appropriate to answer your question and why it is the most appropriate --> 
<!-- List every assumption that your test requires and evaluate it. -->

## Test, results and interpretation
<!-- What are the results of your test? --> 
<!-- What do the results of this test mean? What is the practical significance? --> 