---
title: 'Lab 1 Question 2: Are Democratic voters more enthusiastic about Joe Biden or Kamala Harris?'
author: "Yao Chen, Jenny Conde, Satheesh Joseph, Paco Valdez, Yi Zhang"
output: 
  bookdown::pdf_document2: 
    toc: true
---

\clearpage

```{r load packages, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2) 
library(tidyverse) # if you want more, but still core, toolkit
library(haven)
```


```{r load and clean data, echo=FALSE, warning=TRUE, message=FALSE}
# Load data
rawData <- read_dta('data/anes_timeseries_2020_stata_20210211.dta')

# Useful variables
# V201018 -- PRE: PARTY OF REGISTRATION
# V201151 -- PRE: FEELING THERMOMETER: JOE BIDEN, DEMOCRATIC PRESIDENTIAL CANDIDATE
# V201153 -- PRE: FEELING THERMOMETER: KAMALA HARRIS, DEMOCRATIC VICE-PRESIDENTIAL CANDIDATE
# V201156 -- PRE: Feeling Thermometer: Democratic Party
MaxSentimentScore <- 100
MinSentimentScore <- 0
DemocratPartyCode <- 1

democraticDf <- data.frame(party=rawData$V201018,
                           Biden=rawData$V201151, 
                           Harris = rawData$V201153) %>% 
                filter(party == DemocratPartyCode) %>%
                filter(Biden >= MinSentimentScore & Biden <= MaxSentimentScore) %>%
                filter(Harris >= MinSentimentScore & Harris <= MaxSentimentScore)
```

# Importance and Context
Like many events in the year 2020, the 2020 United States general election was 
unprecedented. The Democrats scored a major victory in the presidential election;
in particular, the election of Kamala Harris as the first female Vice President, 
the first Asian-American Vice President,
and the first African-American Vice President was groundbreaking. 
The new President and Vice President have a long road ahead of them to address
many of today's pressing issues, including the pandemic and economic crisis. However, we 
are curious if Democratic voters had differing levels of enthusiasm for Biden and
Harris leading up to the election. Were Democratic voters more enthusiastic about Biden
leading up to the election or Harris? Knowing the public perception of each individual
could influence their public image during the course of Biden's presidency. 
In this report, we analyze the relative level of support
from Democratic voters between Biden and Harris using the data from the American
National Election Studies (ANES) 2020 Time Series Study and a paired t-test.


# Description of Data

The American National Election Studies (ANES) conducts surveys of voters in the 
United States, with a flagship survey occurring immediately before and after 
each presidential election. ANES survey fits well in our research question. The 
variables tracked in the survey gives a multi-dimensional view about voters.
After careful exploration, we selected the most relevant pre-election variables:

- `V201018` -- `PARTY OF REGISTRATION`

- `V201151` -- `FEELING THERMOMETER: JOE BIDEN, (D) PRESIDENTIAL CANDIDATE`

- `V201153` -- `FEELING THERMOMETER: KAMALA HARRIS, (D) VICE-PRESIDENTIAL CANDIDATE`

that operationalize the concepts in the research question to the best.

We'll use `V201018` to select Democratic voters. We'll use `V201151` and `V201153`
to represent voter enthusiasm for the respective candidates. Voter enthusiasm
is measured on a scale from 0 to 100, with 100 being the most enthusiastic.

As no legal data could identify to whom a party registrant voted for, we assume that the registrant voted for the candidate from their registered party (i.e. A voter registered as a Democrat likely voted for the Democratic candidate, Joe Biden). We considered using the  
`V201156` & `V201157` - FEELING THERMOMETER variables to predict which candidate 
an individual would vote for. However, we felt this would involve too many ambiguous
assumptions, and we preferred using PARTY OF REGISTRATION instead.

For both of the `FEELING THERMOMETER` variables, there were non-answer values,
which we removed from the data. Out of 1861 voters registered as Democrats,
1788 have valid answers for both questions.

The data are summarized below:

``` {r echo=FALSE}
summary(democraticDf)
```

As can be seen, the mean feeling score for both are fairly similar, and because
distributions only represent Democratic voters, 
the scores are both fairly skewed, and each hisogram in Figure 1 shows a left tail.

``` {r echo=FALSE, fig.cap="Histograms of Democrats\' feelings towards Biden and Harris. Since our sample only includes Democrats, it makes sense that the distributions have a left skew.", fig.pos='!h', out.extra='', out.width = "60%", fig.align="center"}
par(mfrow=c(1, 2))
hist(democraticDf$Biden, main="Voter Feelings - Biden", xlab = "Feeling Score", ylab = "Number of people", col='blue')
hist(democraticDf$Harris, main="Voter Feelings - Harris", xlab = "Feeling Score", ylab = "Number of people",  col='red')
```


<!-- Our team also verified the credibility of the participant  -->
<!-- feeling to the Party, by cross-checking the participant loyalty to the Democratic -->
<!-- party using the variable V201156. -->


# Most appropriate test 
As the same voter is expressing her/his feelings towards both candidates and the
feelings are recorded as non-ordinal values,
the correct test is a paired t-test.

We evaluate the assumptions for paired t-test:

1. The dependent variable (feelings rating) is metric since it is on a scale from 0 to 100.

2. Given the sampling frame based on a cross-section of registered addresses across 50 states
and the District of Columbia, we feel the data are sufficiently close to be i.i.d.

3. As can be seen in the histograms in Figure 1, the data is fairly skewed and far from symmetric, so it is a concern here. However given the sample size (1788), the CLT should apply and make the sampling distribution of the mean reasonably normal-like.

#    Test, results and interpretation
For the test itself, we establish the *null hypothesis* to be that the average level of
support of Democrats for Biden ($\mu_B$) and Harris ($\mu_H$) are the same.

And *the alternative hypothesis* is that they're not.

Given we have no strong initial inclination in either direction, this should be a two tailed test.

We'll be using the standard $5\%$ significance level.

$H_0: \mu_B=\mu_H \hspace{4 cm} H_a: \mu_B \neq \mu_H \hspace{4 cm} \alpha=0.05$

```{r Test, echo=FALSE, warning=TRUE, message=FALSE}
t.test(democraticDf$Harris, democraticDf$Biden, paired = TRUE)
```

From the test, we can see that the p-value is $0.0743$, which is greater than the
significance level of $\alpha = 0.05$, therefore we failed to reject the null hypothesis.

However given the p-value is fairly close to $0.05$, it could be considered as a
marginally statistically significant result for the two tailed test. Even though we do not have sufficiently strong evidence to
believe Democratic voters' level of support for Biden and Harris are different given the data
at a $5\%$ significance level, there does seem to be some level of disparity.

In addition, a 95% confidence interval for the difference of voter feelings' mean between Biden and Harris $\mu_B-\mu_H$ is (-1.4904, 0.0698). Because the value 0 is included in this
confidence interval, we cannot reject the null hypothesis at this significance level.
We are 95% confident that this interval contains the true difference in levels of support.

On a practical level, we can see that the mean of the differences is fairly small ($0.71$),
less than 1 point on the $0-100$ scale.

Furthermore, as can be seen in Figure 2 below, when we plot the difference across the board (with 20 bins, each
representing a 5 points difference), we can see that vast majority of Democrat voters' feeling score for
Biden and Harris are close to each other, so the effect size is indeed small.

``` {r fig, fig.cap = "Histogram of absolute differences between individuals\' scores for Biden and for Harris. Most individuals have a small difference in feelings between the two.", fig.align = "center", echo=FALSE, fig.pos='!h', out.extra='', out.width = "60%"}
feeling_diffs = abs(democraticDf$Biden - democraticDf$Harris)

hist(feeling_diffs, breaks=20, main="Voter Feelings - Difference", xlab = "Difference in Feeling Score", ylab = "Number of people")
```

``` {r echo=FALSE}

percentage_equal <- length(feeling_diffs[feeling_diffs == 0]) / length(feeling_diffs)
percentage_within_5 <- length(feeling_diffs[feeling_diffs <= 5]) / length(feeling_diffs)
percentage_within_20 <- length(feeling_diffs[feeling_diffs <= 20]) / length(feeling_diffs)

paste(100 * round(percentage_equal, 4), "% of Democratic voters gave Biden and Harris the same score.")
paste(100 * round(percentage_within_5, 4), "% of Democratic voters gave Biden and Harris scores within 5 pointse.")
paste(100 * round(percentage_within_20, 4), "% of Democratic voters gave Biden and Harris scores within 20 points.")

```


Although curiously, there are a few people who gave Biden and Harris very different
feeling scores, as much as 100!


Although the results from this pre-election survey are not statistically significant,
having this knowledge can establish a baseline for how Democrats view Biden
versus Harris. This could influence how the Democratic Party uses each individual
to influence Party activities. In particular, knowing how the feelings for each
individual changes over time could impact which candidate the Democratic Party runs
in the 2024 election. We now have a baseline for understanding how Democratic
voters feel towards the two individuals, and keeping track of these feelings will
be important going into the future.




