---
title: 'Lab 1: Question 1'
author: "Yao Chen, Jenny Conde, Satheesh Joseph, Paco Valdez, Yi Zhang"
output: pdf_document
---

```{r load packages, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2) 
library(tidyverse) # if you want more, but still core, toolkit
library(haven)
```

```{r load and clean data, echo=FALSE, warning=TRUE, message=FALSE}
# Load data
df <- read_dta('data/anes_timeseries_2020_stata_20210211.dta')

# Pick out the useful variables
# V201018 -- PRE: PARTY OF REGISTRATION
# V201507x -- PRE: SUMMARY: RESPONDENT AGE
df_small <- data.frame(party=df$V201018, age=df$V201507x)

DEMOCRATS <- 1
REPUBLICANS <- 2

# Clean up the data
df_clean <- df_small
df_clean <- df_clean[df_clean$age > 0,] # Remove non-respondent for age
df_clean <- df_clean[(df_clean$party == DEMOCRATS | df_clean$party == REPUBLICANS),] # Only include D or R
```


## Importance and Context
Like many events in the year 2020, the 2020 United States general election was 
unprecedented. Occurring in the midst of a global pandemic with political 
polarization at an all time high and the most diverse candidate pool in United 
States history, the 2020 election posed new challenges and opportunities for 
American citizens and politicians. One key component that both major political 
parties utilized was appealing to voters and encouraging voter turnout. In order 
to appeal to the correct demographic base, it is helpful to understand who 
comprises each political party. One distinguishing factor could be age. In this 
report, we analyze the relative ages of voters registered as either Republican or 
Democratic and we use comprehensive data from the American National Election 
Studies (ANES) 2020 Time Series Study. Understanding age could help politicians
target their campaigns to appropriate demographics and reach audiences with whom
their messages will resonate.

## Description of Data
The ANES data set contains information from 8,280 pre-election interviews with 
U.S. citizens of voting age. Two variables are particularly relevant for us 
to answer this question:

- `V201018`: `PARTY OF REGISTRATION`

- `V201507x`: `SUMMARY: RESPONDENT AGE`

We noticed for both variables, there are irrelevant answers in the data set.
For `PARTY OF REGISTRATION`, we'll only keep Democrats and Republicans, and remove
other parties as well as other non-answers, because we're only interested in the
supporters of these two parties.

Similarly, for `SUMMARY: RESPONDENT AGE`, we will remove people who refused to answer.

After these cleanup operations, we are left with only $3074$ observations to work with.

Looking at their summaries, it looks like the variables are now all in the correct range.

``` {r echo=FALSE}
summary(df_clean)
```

From the following graphs, we can see that the number of Democrats and Republicans are not too disparate and the Age
distribution is not very skewed.

Notice the age number has a hard cutoff at 80 due to the way the survey is constructed, so
everyone above age 80 simply gets grouped into "80 or older" group, so actually the mean
age is somewhat under-representing the true average age of the participants.

``` {r echo=FALSE}
par(mfrow=c(1, 2))
pie(table(df_clean$party), main="Participants Party Afflication", labels=c('Democrats', 'Republicans'))
hist(df_clean$age, main="Participants Age distribution", xlab = "age group", ylab = "Number of people")
```


## Most appropriate test
The unpaired t-test seems to be the most appropriate to answer this question.

1. Even though there are only 3074 valid samples, it is still large enough for the CTE to apply
on the sample average, so it satisfies the normality condition

2. Given the sampling frame based on a cross-section of registered addresses across 50 states
and the District of Columbia, we feel the data are sufficiently close to be i.i.d.

3. And of course `age` is a metric scale variable.

Furthermore, because we're comparing two distinct groups of people with no natural
pairing between them, this directs us to the unpaired t-test.

## Test, results and interpretation
For the test itself, we establish the *null hypothesis* to be that the average age
of Democrats and Republicans are the same.

And *the alternative hypothesis* is that they're not.
Given there is no bias towards either side, this should be a two tailed test.

We'll be using the standard $5%$ significance level.

``` {r}
t.test(df_clean$age ~ df_clean$party)
```

From the test, it looks like we have a very very small p-value, representing a highly
significant result.

This gives us evidence to reject the null hypothesis in favor of the alternative and
believe that the average age of the Democrats and the Republicans are indeed different,
given the data and a 5% significance level.

Practically, as can be seen above, the average age of a Republican is more than 3 years
older than a Democrat.

Plotting the distribution of ages within each group, it's clear that Democrats
participants are more evenly distributed between "young" and "old" whereas Republicans
are much more skewed towards people above 60.

At the same time, there are larger proportion of Democrats in the entire age group
between 18 and 50 than Republicans and vice versa for over 50.

```{r, echo = FALSE, warning=FALSE, message=FALSE}
# Add categorical variable indicating Democratic or Republican
df_clean <- mutate(df_clean, party_affiliation = ifelse(party == DEMOCRATS, "Democratic Party", "Republican Party"))

# Find average age in each political party
av_ages <- df_clean %>%
  group_by(party_affiliation) %>%
  summarize(av_age = mean(age))

# Create density plot showing age distributions of the two parties
df_clean %>%
  ggplot() +
  aes(x = age, color = party_affiliation) +
  geom_density() + 
  geom_vline(data = av_ages, aes(xintercept=av_age, 
                                 color = party_affiliation),
             linetype = "dashed") +
  scale_color_manual(values = c("blue2", "red2")) + 
  labs(title = "Distribution of ages across political parties",
       caption = "The dotted lines represent the average age in each respective political party.") +
  theme_minimal()
```

More quantitatively, if we look at the correlation between the two groups, we see that
there is a mild correlation, meaning that there is mildly strong (linear) relationship
between the political party affiliation and age.

``` {r}
cor(df_clean$party, df_clean$age)
```

