---
title: 'Lab 1 Question 3: Are survey respondents who have had someone in their home infected by COVID-19 more likely to disapprove of the way their governor is handling the pandemic?'
author: "Yao Chen, Jenny Conde, Satheesh Joseph, Paco Valdez, Yi Zhang"
output: 
  bookdown::pdf_document2: 
    toc: true
---

```{r load packages, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2) 
# library(tidyverse) # if you want more, but still core, toolkit
library(haven)

library(tidyverse) 
library(magrittr)
library(knitr)
library(patchwork)
theme_set(theme_bw())
options(tinytex.verbose = TRUE)
```

```{r load and clean data, echo=FALSE, warning=TRUE, message=FALSE}
# Load data
df <- read_dta('data/anes_timeseries_2020_stata_20210211.dta')

# Useful variables
# V201145 -- PRE: APPROVE OR DISAPPROVE R’S GOVERNOR HANDLING COVID-19
  # 1 = approve, 2 = disapprove
gov_approve = 1
gov_disapprove = 2
# V201624 -- PRE: ANYONE IN HOUSEHOLD TESTED POS FOR COVID-19
  # 1 = someone tested positive, 2 = no one tested positive 
pos_test = 1
no_pos_test = 2
# Less useful but still interesting
# V201146 -- PRE: HOW MUCH APPROVE/DISAPPROVE R’S GOVERNOR HANDLING COVID-19
# V201625 -- PRE: ANYONE IN HOUSEHOLD COVID-19 BASED ON SYMPTOMS

df_small <- data.frame(gov=df$V201145, covid=df$V201624)

df_clean <- df_small
df_clean <- df_clean[df_clean$covid > 0,] # remove interview breakoff/refused
df_clean <- df_clean[df_clean$gov > 0,] # remove don't know/refused

```


## Importance and Context
The COVID-19 has changed everything, and one of the most important aspects is how the campaigns for the US 2020 elections were conducted. The pandemic forced to cancel conventions and prompted many states to change how people get and submit their ballots, with all the uncertainty it created. And to make things worse, all of this happened in the middle of social distancing restrictions not seen since almost a century. It's essential to understand how elected officials' approval changed when people got in close contact with COVID-19.

## Description of Data
The ANES data set contains information from 8,280 pre-election interviews with 
U.S. citizens of voting age. Two variables are particularly relevant for us 
to answer this question:

- `V201145`: `APPROVE OR DISAPPROVE R’S GOVERNOR HANDLING COVID-19`

- `V201624`: `ANYONE IN HOUSEHOLD TESTED POS FOR COVID-19`

```{r}
summary(df_clean)
```


```{r, echo = FALSE, fig.cap='Voters\' Approval of Governor and Household COVID-19 Tests', fig.pos='!h'}
df_clean_mod <- mutate(df_clean, gov_approval = ifelse(gov == gov_approve, "Approve", "Disapprove"))
df_clean_mod <- mutate(df_clean_mod, positive_test = ifelse(covid == pos_test, "Positive Test", "No Positive Test"))

gov_count <- df_clean_mod %>%
  group_by(gov_approval) %>%
  summarize(count=n())

pos_test_count <- df_clean_mod %>%
  group_by(positive_test) %>%
  summarize(count = n())

gov_pie_chart <- gov_count %>%
  ggplot() +
  aes(x = "", y = count, fill = gov_approval) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y", start = 0) +
  labs(fill = "Approve or Disapprove of \nGovernor's Response",
       title = "Distribution of Approval of\nGovernor's COVID-19 Handling") +
  theme_void()

covid_pie_chart <- pos_test_count %>%
  ggplot() +
  aes(x = "", y = count, fill = positive_test) + 
  geom_bar(stat = "identity", width = 1) + 
  coord_polar("y", start = 0) +
  labs(fill = "COVID-19 Case in Household",
       title = "Distribution of Households\nwith Positive COVID-19 Tests") +
  theme_void()

(gov_pie_chart) | (covid_pie_chart)
```


```{r, echo = FALSE, fig.pos = '!h'}
summary_table <- prop.table(table(df_clean_mod %>% select(3, 4)))
kable(summary_table,
      digits = 3,
      caption = "Cross Tab of Governor Approval and Positive COVID-19 Tests",
      booktabs = TRUE)
```


## Most appropriate test 
<!-- Explain which test is the most appropriate to answer your question and why it is the most appropriate --> 
<!-- List every assumption that your test requires and evaluate it. -->
Wilcoxon rank test


## Test, results and interpretation
<!-- What are the results of your test? --> 
<!-- What do the results of this test mean? What is the practical significance? --> 

```{r}

table(df_clean$gov, as.numeric(df_clean$gov))
df_clean$covid_positive = df_clean$covid == pos_test
summary(df_clean$covid_positive)

wilcox.test(as.numeric(df_clean$gov)~df_clean$covid_positive)
```


```{r}
cor.test(as.numeric(df_clean$gov), as.numeric(df_clean$covid_positive), method='spearman')
```