---
title: 'Lab 1 Question 3: Are survey respondents who have had someone in their home infected by COVID-19 more likely to disapprove of the way their governor is handling the pandemic?'
author: "Yao Chen, Jenny Conde, Satheesh Joseph, Paco Valdez, Yi Zhang"
output: 
  bookdown::pdf_document2: 
    toc: true
---

\clearpage

```{r load packages, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2) 
library(ggcorrplot)
library(haven)
# install.packages("rstatix")
# install.packages('coin')
library(rstatix)

library(tidyverse) 
library(magrittr)
library(knitr)
library(patchwork)
theme_set(theme_bw())
options(tinytex.verbose = TRUE)
```

```{r load and clean data, echo=FALSE, warning=TRUE, message=FALSE}
# Load data
df <- read_dta('data/anes_timeseries_2020_stata_20210211.dta')

# Useful variables
# V201145 -- PRE: APPROVE OR DISAPPROVE R’S GOVERNOR HANDLING COVID-19
  # 1 = approve, 2 = disapprove
gov_approve = 1
gov_disapprove = 2
# V201624 -- PRE: ANYONE IN HOUSEHOLD TESTED POS FOR COVID-19
  # 1 = someone tested positive, 2 = no one tested positive 
pos_test = 1
no_pos_test = 2
# Less useful but still interesting
# V201146 -- PRE: HOW MUCH APPROVE/DISAPPROVE R’S GOVERNOR HANDLING COVID-19
# V201625 -- PRE: ANYONE IN HOUSEHOLD COVID-19 BASED ON SYMPTOMS

df_small <- data.frame(gov=df$V201145, covid=df$V201624)

df_clean <- df_small
df_clean <- df_clean[df_clean$covid > 0,] # remove interview breakoff/refused
df_clean <- df_clean[df_clean$gov > 0,] # remove don't know/refused
```


# Importance and Context
The COVID-19 pandemic has caused significant upheaval in the lives
of U.S. citizens. The pandemic has also become highly politicized, with each political
party criticizing the other for its response (or lack thereof) to the pandemic,
including stay at home orders, closures of local businesses, and the encouragement
of masks and social distancing.
Simultaneously, U.S. citizens have been forming their own opinions about local
and federal responses, and in this report we analyze if citizens' opinions of 
their elected governors' response to the pandemic differ depending on whether 
or not someone in their household
has tested positive for COVID-19. Knowing this information prior to the 2020 election
can help state governments and political parties understand the opinions of 
their constituents, which could influence their campaign strategies. To test if
the approval rate of governors is different between households that have tested
positive for COVID-19 versus those that have not, we utilize comprehensive data 
from the American National Election Studies (ANES) 2020 Time Series Study and a 
Wilcoxon rank sum test.

<!--The COVID-19 has changed everything, and one of the most important aspects is how the campaigns for the US 2020 elections were conducted. The pandemic forced to cancel conventions and prompted many states to change how people get and submit their ballots, with all the uncertainty it created. And to make things worse, all of this happened in the middle of social distancing restrictions not seen since almost a century. It's essential to understand how elected officials' approval changed when people got in close contact with COVID-19.-->

# Description of Data
The ANES data set contains information from 8,280 pre-election interviews with 
U.S. citizens of voting age. Two variables are particularly relevant for us 
to answer this question:

- `V201145`: `APPROVE OR DISAPPROVE R’S GOVERNOR HANDLING COVID-19`

- `V201624`: `ANYONE IN HOUSEHOLD TESTED POS FOR COVID-19`

While there are other variables in the ANES data set that could have been helpful
in answering this question, including 
`V201146 -- PRE: HOW MUCH APPROVE/DISAPPROVE R’S GOVERNOR HANDLING COVID-19`, 
we believe the two highlighted above best address and most successfully 
operationalize our research question.

Both variables are binary variables. In `V201145`, governor approval is 
represented by the value 1, while disapproval is represented by the number 2. We
remove the remaining irrelevant values, including those for refusal to answer and
interviewees whose responses got cut off. The breakoff of a survey response 
could happen for a variety of reasons, including lack of Internet or phone service.

For `V201624`, the value 1 is assigned if there has been a positive
COVID-19 test in the respondent's household; the value 2 is assigned if there has
been no positive COVID-19 test in the respondent's household. We remove the 
irrelevant values corresponding to those who refused to answer.

After cleaning the data, we are left with 8,138 valid observations. We can see
from the data summary below that our variables are in the expected ranges.

```{r}
summary(df_clean)
```

We additionally do some initial data exploration. Table 1 shows how our data is
divided among our two Boolean variables. It is clear that the vast majority of
households have not had a positive COVID-19 test. However, further analysis about
how this relates to governor approval is less clear. Figure 1 helps visualize 
how the data is split. The pie chart on the left shows the approval/disapproval
rates for the households that have received a positive COVID-19 test, and the pie
chart on the left is for households that have not had a positive test. We can see
that there is some difference in how the approval and disapproval rates are split--
on average, 44% of households that have tested positive disapprove of their governor's
response, while this value is only 38% for households with no positive COVID-19 test.
We want to test if there is a statistical significance in this difference, and our
methods are described further in the next section.

```{r, echo = FALSE, fig.cap='Distribution of Voters\' Approval of Governor based on Positive COVID-19 Test in Household', fig.pos='!h', out.extra='trim={0 3cm 0 3cm},clip', fig.align="center"}
df_clean_mod <- mutate(df_clean, gov_approval = ifelse(gov == gov_approve, "Approve", "Disapprove"))
df_clean_mod <- mutate(df_clean_mod, positive_test = ifelse(covid == pos_test, "Positive Test", "No Positive Test"))

# gov_count <- df_clean_mod %>%
#   group_by(gov_approval) %>%
#   summarize(count=n())
# 
# pos_test_count <- df_clean_mod %>%
#   group_by(positive_test) %>%
#   summarize(count = n())

# df_clean_mod

gov_given_pos_test <- df_clean_mod[df_clean_mod$positive_test == "Positive Test",] %>%
  group_by(gov_approval) %>%
  summarize(count = n())

gov_given_no_pos_test <- df_clean_mod[df_clean_mod$positive_test == "No Positive Test",] %>%
  group_by(gov_approval) %>%
  summarize(count = n())

pie_pos_test <- gov_given_pos_test %>%
  ggplot() +
  aes(x = "", y = count, fill = gov_approval) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  labs(fill = "Approve or Disapprove of \nGovernor's Response",
       title = paste("Positive COVID-19 Test\nin Household, n =", 
                     sum(gov_given_pos_test$count))) +
  theme_void() + 
  scale_fill_manual(values = c("turquoise3", "tomato1")) +
  geom_text(aes(label = paste(round(count/sum(count)*100), "%")),
            position = position_stack(vjust=0.5))

pie_no_pos_test <- gov_given_no_pos_test %>%
  ggplot() +
  aes(x = "", y = count, fill = gov_approval) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  labs(fill = "Approve or Disapprove of \nGovernor's Response",
       title = paste("No Positive COVID-19\nTest in Household, n =", 
                     sum(gov_given_no_pos_test$count))) +
  theme_void() +
  scale_fill_manual(values = c("turquoise3", "tomato1")) +
  geom_text(aes(label = paste(round(count/sum(count)*100), "%")),
            position = position_stack(vjust=0.5))

# gov_pie_chart <- gov_count %>%
#   ggplot() +
#   aes(x = "", y = count, fill = gov_approval) +
#   geom_bar(stat = "identity", width = 1) +
#   coord_polar("y", start = 0) +
#   labs(fill = "Approve or Disapprove of \nGovernor's Response",
#        title = "Distribution of Approval of\nGovernor's COVID-19 Handling") +
#   theme_void()

# covid_pie_chart <- pos_test_count %>%
#   ggplot() +
#   aes(x = "", y = count, fill = positive_test) + 
#   geom_bar(stat = "identity", width = 1) + 
#   coord_polar("y", start = 0) +
#   labs(fill = "COVID-19 Case in Household",
#        title = "Distribution of Households\nwith Positive COVID-19 Tests") +
#   theme_void()

# (gov_pie_chart) | (covid_pie_chart)
(pie_pos_test) | (pie_no_pos_test)
```


```{r, echo = FALSE, fig.pos = '!h'}
summary_table <- prop.table(table(df_clean_mod %>% select(3, 4)))
kable(summary_table,
      digits = 3,
      caption = "Cross Tab of Governor Approval and Positive COVID-19 Tests",
      booktabs = TRUE)
```



# Most appropriate test 
To answer our question of whether individuals' opinions of their governor's response
to the COVID-19 pandemic depends on whether someone in their household has tested
positive for COVID-19, we use a Wilcoxon Rank Sum Test. We believe this test is
most appropriate for the given data because our data is ordinal: approval/disapproval
of the governor's response is a categorical variable with a natural ordering. A
parametric test based on some underlying distribution that resembles a normal 
distribution would not be appropriate. Furthermore, the two groups of interest
(people who had a positive test in their household v.s.
people who did not) are distinct observations, and they don't have a natural pairing.

There are two main assumptions for the Wilcoxon Rank Sum Test:

1. Ordinal variable: As discusses already, approval/disapproval of the governor's
response is an ordinal variable.

2. i.i.d.: Given the sampling frame based on a cross-section of registered 
addresses across 50 states
and the District of Columbia, we feel the data are sufficiently close to be i.i.d.

With all assumptions sufficiently met, we proceed with a Wilcoxon Rank Sum Test.

# Test, results and interpretation
<!-- What are the results of your test? --> 
<!-- What do the results of this test mean? What is the practical significance? --> 
We establish the *null hypothesis* to be that there is no difference in the opinions
of the governor's response between individuals who have had a positive COVID-19 test in their household compared to those who have not had a positive test. Our *alternative hypothesis* is
that there is a difference between the opinions of the two groups.
Given we have no strong initial inclination in either direction, we conduct a two tailed test. We use the standard $5\%$ significance level.

``` {r echo=FALSE}
df_clean$approval <- recode(as.character(df_clean$gov), '1' = 'Approve', '2' = 'Disapprove')
df_clean$covid_tested <- recode(as.character(df_clean$covid), '1' = 'Positive', '2' = 'Not positive')
```

<!--First take a look at the table in each of the 4 cases for context, then run the test.-->
```{r, echo = FALSE}
table_results <- table(df_clean$approval, df_clean$covid_tested)
```

```{r}
wilcox.test(df_clean$gov ~ df_clean$covid)
```


From the test we can see that the p-value is $0.0377$, which is less than the significance
level $\alpha = 0.05$.We reject the null hypothesis in favor of the
alternative that people with a positive COVID-19 test in their household do have a different
opinion of their governor's response than people without.

There is little practical significance in these results. Because we have 
a large sample size, we calculated the Wilcoxon Rank Sum effect size. We found
the effect size to be equal to $0.023$, which is a very low value and indicates a very 
low-magnitude effect. Practically speaking the difference between the two groups in 
our test is very minimal, and our pie charts in Figure 1 further corroborate 
this finding. Interestingly, however, it is likely that politicians and political
parties made specific appeals in their campaigns to families who have received
a positive COVID-19 test in hopes of winning their vote due to their dissatisfaction
with their governor's or state's response to COVID-19. Even this small difference
likely caused a change in campaign strategies.

```{r echo=FALSE, fig.cap='Correlation Table for Voters\' Approval of Governor and Household COVID-19 Tests', fig.pos='!h'}
r <- cor(data.frame('Approve' = as.numeric(df_clean_mod$gov==1), 'Disapprove' = as.numeric(df_clean_mod$gov==2), 'Positive' = as.numeric(df_clean_mod$covid==1), 'Not_Positive' = as.numeric(df_clean_mod$covid==2)))
plt <- ggcorrplot(r[3:4,1:2], lab = TRUE, ggtheme = ggplot2::theme_dark(), digits = 3)
```

<!-- As presented in Fig. 2, having a positive test in the household does linearly 
correlate to less likely to approve the Governor's handling of the pandemic.
However given the magnitude of the the correlation coefficient. The linear relationship isn't
very strong. -->

```{r, echo = FALSE}
eff_size <- df_clean %>% wilcox_effsize(gov ~ covid)
effect <- eff_size$effsize
```

